\begin{abstract}
Vector-quantized autoencoders (VQ-VAEs) underpin much of modern discrete
representation learning, from tokenizers for autoregressive image
generation~\citep{esser2021vqgan,yu2022vitvqgan,sun2024llamagen} to the
latent spaces of high-resolution diffusion
models~\citep{rombach2022ldm}. Their training, however, relies on the
straight-through estimator
(STE,~\citealp{bengio2013ste,vandenoord2017vqvae}), which copies the
upstream gradient through the non-differentiable nearest-neighbour
lookup as if the quantization were the identity. This bypasses the
geometric relationship between the encoder output $\ze$ and its assigned
codebook vector $\qvec$, and is widely blamed for codebook collapse and
the persistent gap between reconstructions produced by codebook-based
and continuous
latents~\citep{huh2023straighteningste,mentzer2024fsq}. We propose
\ours{}, a simple drop-in replacement for STE that routes the backward
pass through the rotation-plus-rescaling transformation that
analytically maps $\ze$ onto $\qvec$. Concretely, we write the quantized
output as $\sg{s\rmat}\,\ze$ where $\rmat$ is a rank-one Householder
reflection aligning $\ze/\|\ze\|$ with $\qvec/\|\qvec\|$ and
$s=\|\qvec\|/\|\ze\|$ is a scalar rescaling, with the stop-gradient
operator applied so that the forward value is numerically identical to
the standard VQ output. The resulting Jacobian of the quantizer is not
the identity but the very rotation-scaling that the forward pass
implicitly performs; this preserves angular information between
$\ze$ and $\qvec$ during backpropagation. The transform costs
$\mathcal{O}(d)$ per latent token in both time and memory because
$\rmat$ is never materialised. We present a rigorous experimental
comparison on CIFAR-10 against VQ-VAE with STE, FSQ, and
Gumbel-Softmax VQ, and a component-wise ablation that separates the
Householder rotation from the scalar rescaling. Our main empirical
finding is cautionary: routing the full $s\cdot\rmat$ Jacobian
naively through the standard VQ-VAE training recipe induces a
sharp codebook collapse after $\sim 1{,}000$ steps, which we trace
to a positive-feedback interaction with the commitment loss. An
ablation localises the failure mode to one specific component of
the estimator and identifies a stable single-component variant.
We release a reference implementation, the exact experimental
configurations, and training logs for all reported runs.
\end{abstract}
